---
title: "Hands-On Exercise 1:</br>A Layered Grammar of Graphics: ggplot2 methods"
author: "Chua Wee Tian"
date: "April 11, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

## 1.1 Learning Outcome

In this chapter, you will learn the basic principles and essential components of ggplot2. At the same time, you will gain hands-on experience on using these components to plot statistical graphics based on the principle of Layered Grammar of Graphics. By then end of this chapter you will be able to apply the essential graphical elements provided by ggplot2 to create elegant and yet functional statistical graphics.

## 1.2 Getting started

### 1.2.1 Installing and loading the required libraries

Before we get started, it is important for us to ensure that the required R packages have been installed. If yes, we will load the R packages. If they have yet to be installed, we will install the R packages and load them onto R environment.

The code chunk on the right assumes that you already have [**pacman**](https://cran.r-project.org/web/packages/pacman/index.html) package installed. If not, please go ahead install pacman first.

```{r}
pacman::p_load(tidyverse)
```

### Import Data

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

### Simple Bar Chart

```{r}
hist(exam_data$MATHS)
```

### ggplot2</br>

-   **Data**: The dataset being plotted.

-   **Aesthetics** take attributes of the data and use them to influence visual characteristics, such as position, colours, size, shape, or transparency.

-   **Geometrics**: The visual elements used for our data, such as point, bar or line.

-   **Facets** split the data into subsets to create multiple variations of the same graph (paneling, multiple plots).

-   **Statistics**, statiscal transformations that summarise data (e.g. mean, confidence intervals).

-   **Coordinate systems** define the plane on which data are mapped on the graphic.

-   **Themes** modify all non-data components of a plot, such as main title, sub-title, y-aixs title, or legend background.

#### geom_bar

```{r}
ggplot(data = exam_data, 
       aes(x = RACE)) +
    geom_bar()
```

#### geom_dotplot

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(dotsize = 0.5)

ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(binwidth=2.5,         
               dotsize = 0.5) +      
  scale_y_continuous(NULL,           
                     breaks = NULL)  
```

#### geom_histogram

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_histogram()       

ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20,            
                 color="black",      
                 fill="light blue")  

ggplot(data=exam_data, 
       aes(x= MATHS, 
           fill = GENDER)) +
  geom_histogram(bins=20, 
                 color="grey30")
```

```{r}
```

```{r}
```
