---
title: "Take-Home Exercise 2: Mini-Challenge 2:</br>Investigating the Anomalous SaidIt Post"
author: "CHUA Wee Tian"
date: "June 7, 2026"
date-modified: "last-modified"
format:
  html:
    toc: true
    code-fold: true
    self-contained: false
execute:
  eval: true
  echo: true
  freeze: true
  warning: false
  message: false
---

# 1. Introduction

## Background

On 17 May 2046 at 4:21am, an anomalous post was published on SaidIt by John Windward. The post appeared to contain gibberish and was inconsistent with normal business operations.

The objective of this investigation is to construct a visual analytics system capable of:

-   Understanding the multi-agent ecosystem
-   Reconstructing the chain of events leading to the anomalous post
-   Identifying the origin of the post contents
-   Detecting historical occurrences of similar behavior
-   Recommending an effective intervention point

## Load the R Package

We use `pacman::p_load()` function to load the required R packages into our working environment.

```{r}
pacman::p_load(tidyverse,jsonlite,lubridate,tidygraph,ggraph,igraph,plotly,scales,visNetwork,DT,tidytext,DiagrammeR)
```

The loaded packages are:

-   **tidyverse** : provide a consistent and coherent approach to data manipulation and analysis
-   **jsonlite** : for working with JSON (JavaScript Object Notation) data
-   **lubridate** : provides functions for working with dates and times
-   **tidygraph** : provides a tidy and consistent way to work with graph data structure
-   **ggraph** : creates visualizations of graphs using the grammar of graphics approach
-   **igraph** : provides functions for creating, analyzing, and visualizing graphs
-   **plotly** : for creating interactive web-based graphs
-   **visNetwork** : creates interactive network visualizations
-   **DT** : creates interactive tables using the DataTables JavaScript library
-   **tidytext** : focuses on text mining and natural language processing (NLP) tasks
-   **DiagrammeR:** package is the standard for generating flowcharts

# 2. Data Preparation

## Import the Data

```{r}
mc2 <- fromJSON("data/MC2 data.json", flatten = TRUE)
org <- fromJSON("data/org_chart.json", flatten = TRUE)

events <- mc2$events %>%
  as_tibble()

head(events)

nodes <- as_tibble(org$nodes)
edges <- as_tibble(org$edges)

head(nodes)
```

## Data Preparation

The **when** column is converted to datetime object using as_datetime().

```{r}
events <- events %>%
  mutate(
    datetime = as_datetime(when)
  )

events %>%
  select(datetime) %>%
  head(6)
```

# 3. Organizational Structure

To understand the overall multi-agent ecosystem and communication pathways.

```{r}
graph_org <- tbl_graph(
  nodes = nodes,
  edges = edges,
  directed = TRUE
)

nodes_df <- graph_org %>% 
  activate(nodes) %>% 
  as_tibble() %>% 
  mutate(
    id = row_number(),         
    group = type               
  )

edges_df <- graph_org %>% 
  activate(edges) %>% 
  as_tibble()                   

visNetwork(nodes_df, edges_df) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visOptions(
    highlightNearest = TRUE,
    selectedBy = "group"        
  ) %>%
  visLegend(
    position = "right",
    main = "Legend"
  ) %>%
  visLayout(randomSeed = 123)
```

The organization operates through a hierarchy of:

-   Company (1)
-   Department (6)
-   Team (19)
-   Person (49)

# 4. Event Activity Overview

To understand the baseline of the types of events logged and frequency.

```{r}
events %>%
  count(short_name) %>%
  ggplot(aes(
    reorder(short_name,n),
    n
  )) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title="Frequency of Event Type",
    x="Event Type",
    y="Count"
  )

event_counts <- events %>%
  count(short_name, sort = TRUE) %>%
  rename(`Event Type` = short_name, Count = n) %>%
  mutate(`% of Total` = percent(Count / sum(Count), accuracy = 0.1))

datatable(event_counts)
```

::: callout-note
-   There are 24 event types recorded in the dataset.
-   We could identify saidit counts are low, which might camouflage anomalies.
:::

# 5. Investigating John Windward

## Extract Relevant Events

```{r}
john_events <- events %>%
  filter(
    map_lgl(parties, ~ any(str_detect(.x, "john_windward"))) & 
    datetime < as.Date("2046-05-18")
  ) %>%
  select(where(~ !all(is.na(.x) | as.character(.x) == ""))) %>%
  arrange(datetime)

datatable(john_events, options = list(pageLength = 10, scrollX = TRUE))
```

## Timeline Reconstruction

```{r}
john_events %>%
  filter(datetime < as.Date("2046-05-18")) %>%
  plot_ly(
    x = ~datetime,
    y = ~short_name,
    type = 'scatter',
    mode = 'markers+lines',
    text = ~id
  ) %>%
  layout(
    yaxis = list(title = "Event Type"),
    xaxis = list(title = "Date/Time")
  )
```

## Interpretation

The timeline reconstructs the exact sequence of actions involving John Windward and associated agent.

On Event Type = saidit_post, there are 3 content_source that points to external .txt files worth investigation:

-   **SwiftWren.txt**
-   **HiddenOrca.txt**
-   **MellowOtter.txt**

The anomalous posts were injected directly from an external file, bypassing normal agent-generated content.

On May 17 there is a sequence of events of

-   queue_subordinate_task
-   saidit_post_check
-   saidit_post
-   delete_file
-   delete_file

All these events happened within the same minute which then suggested coordinated events action.

The two delete_file events happened immediately after the anomalous post which points to a cover up attempt.

# 6. Agent Interaction Network

John Windward is highlighted for ease of locating him within the company and he is the department lead for the Customer Support department.

```{r}
nodes_df <- graph_org %>% 
  activate(nodes) %>% 
  as_tibble() %>% 
  mutate(
    id = row_number(),         
    group = type,
    color.border = ifelse(label == "John Windward", "red", NA),
    borderWidth = ifelse(label == "John Windward", 3, 1)
  )

edges_df <- graph_org %>% 
  activate(edges) %>% 
  as_tibble()                   


visNetwork(nodes_df, edges_df) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visOptions(
    highlightNearest = TRUE,
    selectedBy = list(
      variable = "label",       
      style = 'width: 200px; height: 30px;' 
    )
  ) %>%
  visLegend(
    position = "right",
    main = "Legend",
    useGroups = TRUE,  
    addNodes = list(   
      list(
        label = "John Windward", 
        color = list(background = "lightblue", border = "red"),
        borderWidth = 3
      )
    )
  ) %>%
  visLayout(randomSeed = 123)
```

The organization operates through a hierarchy of:

-   Company (1)
-   Department (6)
-   Team (19)
-   Person (49)

# 7. Tracing the Events

## Suspicious Sequence - “SwiftWren”

Rearranging the events that contain the keyword **"SwiftWren"** and filtered the required columns and display by datetime, we can see the chain of sequence leading to the posting and deleting of the file.

```{r}
Suspicious <- events %>%
  filter(
    #as.Date(datetime) == as.Date("2046-05-17") &
    if_any(everything(), ~ str_detect(as.character(.x), regex("SwiftWren", ignore_case = TRUE)))
  ) %>%
  select(where(~ !all(is.na(.x) | as.character(.x) == ""))) %>%
  select(-when,-details.calendar_emails_sent,-details.contacts,-details.combo,-details.meeting.participants,-details.a2a.payload.participants,-details.details.participants,-details.details.contacts,-details.details.combo) %>% 
  select(datetime, everything()) %>%
  arrange(datetime)

datatable(Suspicious, options = list(pageLength = 10, scrollX = TRUE))
```

### Interpretation:

-   **Emma Harbor (CFO) created the "SwiftWren.txt"**

-   **There was no reading of the file before the post which indicate no verification.**

-   **John Windward execute the post "SwiftWren.txt" and deleted the file.**

-   **Clearly there is something not right in this sequence.**

## Suspicious Sequence - “HiddenOrca”

Rearranging the events that contain the keyword **"HiddenOrca"** and filtered the required columns and display by datetime, we can see the chain of sequence leading to the posting and deleting of the file.

```{r}
Suspicious <- events %>%
  filter(
    #as.Date(datetime) == as.Date("2046-05-17") &
    if_any(everything(), ~ str_detect(as.character(.x), regex("HiddenOrca", ignore_case = TRUE)))
  ) %>%
  select(where(~ !all(is.na(.x) | as.character(.x) == ""))) %>%
  select(-when,-details.calendar_emails_sent,-details.contacts,-details.combo,-details.meeting.participants,-details.a2a.payload.participants,-details.details.participants,-details.details.contacts,-details.details.combo) %>% 
  select(datetime, everything()) %>%
  arrange(datetime)

datatable(Suspicious, options = list(pageLength = 10, scrollX = TRUE))
```

### Interpretation:

-   **We could not identify who created the "HiddenOrca.txt"**

-   **We might have to retrieve earlier logs for further investigation.**

-   **There was no reading of the file before the post which indicate no verification.**

-   **John Windward execute the post "HiddenOrca.txt" and deleted the file.**

## Suspicious Sequence - “MellowOtter”

Rearranging the events that contain the keyword **"MellowOtter"** and filtered the required columns and display by datetime, we can see the chain of sequence leading to the posting and deleting of the file.

```{r}
Suspicious <- events %>%
  filter(
    #as.Date(datetime) == as.Date("2046-05-17") &
    if_any(everything(), ~ str_detect(as.character(.x), regex("MellowOtter", ignore_case = TRUE)))
  ) %>%
  select(where(~ !all(is.na(.x) | as.character(.x) == ""))) %>%
  select(-when,-details.calendar_emails_sent,-details.contacts,-details.combo,-details.meeting.participants,-details.a2a.payload.participants,-details.details.participants,-details.details.contacts,-details.details.combo) %>% 
  select(datetime, everything()) %>%
  arrange(datetime)

datatable(Suspicious, options = list(pageLength = 10, scrollX = TRUE))
```

### Interpretation:

-   **Noah Mariner (COO) created the "MellowOtter.txt"**

-   **There was no reading of the file before the post which indicate no verification.**

-   **John Windward execute the post "MellowOtter.txt" and deleted the file.**

-   **Clearly there is something not right in this sequence.**

# STOP HERE\>\>\>\>\>\>\>\>\>\>\>\>\>\>\>\>\>\>\>\>\>\>\>\>\>\>\>\>\>\>\>\>\>\>\>\>\>\>\>\>\>\>\>\>\>\>\>\>\>

# 11. Intervention Analysis

## Proposed Control Point

The strongest intervention point occurs before:

``` text
queue_subordinate_task
        →
read_file
        →
saidit_post
```

## Why This Point?

1.  All anomalous chains pass through it
2.  Minimal impact on normal operations
3.  High prevention coverage

## Intervention Diagram

```{r}


grViz("
digraph {

queue_task -> validation

validation -> read_file

read_file -> saidit_post

}
")
```

## Justification

A validation layer between task assignment and file execution maximises anomaly prevention while minimising disruption.

# 12. Conclusion

This investigation demonstrates that the anomalous SaidIt post was not an isolated malfunction. Visual analytics reveals recurring patterns involving automated task assignment, file access, and external content propagation.

Historical evidence suggests the existence of a systemic vulnerability capable of generating repeated anomalous behavior across multiple agents.

A validation checkpoint placed between task assignment and file execution would likely prevent future occurrences while preserving normal operational workflows.

# Conclusion

The investigation demonstrates that the anomalous SaidIt behaviour can be explained through agent-task propagation, file access behaviour and external content execution pathways. Historical patterns indicate systemic risk rather than isolated user error. A validation checkpoint is recommended.

#Final

The findings suggest that the SaidIt post was not an isolated malfunction but part of a broader propagation pattern involving HiddenOrca, MellowOtter and SwiftWren instruction chains.
