---
title: "Visual Analytics for Data Discovery"
subtitle: "Take-Home Exercise 1: Executive Summary"
author: "Chua Wee Tian"
date: "May 24, 2026"
format:
  revealjs:
    theme: simple
    slide-number: true
    incremental: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
pacman::p_load(
  tidyverse, ggrepel, ggthemes, ggridges, ggdist, colorspace, patchwork, scales
)
```

```{r}
insurance <- read_delim("data/Dataset of motor insurance portfolio.csv")
```

```{r}
insurance <- insurance |>
  mutate(
    fuel_type = replace_na(fuel_type, "Unknown"),
    vehicle_value = replace_na(
      vehicle_value,
      median(vehicle_value, na.rm = TRUE)
    ),
    vehicle_age = replace_na(
      vehicle_age,
      median(vehicle_age, na.rm = TRUE)
    ),
    age_driving_licence = replace_na(
      age_driving_licence,
      median(age_driving_licence, na.rm = TRUE)
    ),
    fire_incurred = replace_na(fire_incurred, 0),
    occupants_incurred = replace_na(
      occupants_incurred,
      0
    )
  )
```

```{r}
insurance <- insurance |>
  mutate(
    total_claims = as.numeric(total_claims),
    total_exposure = as.numeric(total_exposure),
    total_incurred = as.numeric(total_incurred),
    total_premium = as.numeric(total_premium),

    loss_ratio = if_else(
      total_premium > 0,
      total_incurred / total_premium,
      0
    ),

    profitability_band = case_when(
      loss_ratio < 0.6 ~ "Profitable",
      loss_ratio < 1 ~ "Marginal",
      TRUE ~ "Unprofitable"
    ),

    age_band = case_when(
      driver_age < 30 ~ "Young",
      driver_age < 60 ~ "Middle",
      TRUE ~ "Senior"
    )
  )
```

## Introduction and Objectives

-   Investigate profitability drivers within the insurance portfolio
-   Identify factors influencing claims volatility
-   Explore demographic and vehicle related risk patterns
-   Apply visual analytics techniques for business interpretation
-   Support underwriting and pricing decisions

::: notes
The objective of this investigation was to understand how underwriting variables influence profitability and claims behaviour within the motor insurance portfolio. The study applied visual analytics techniques to uncover hidden relationships between policy characteristics, demographic variables and claims outcomes. Using both univariate and bivariate visualisations, the findings provide actionable insights to support data driven underwriting, pricing optimisation and portfolio management.
:::

## Data Preparation Process

-   Managed missing and inconsistent values
-   Converted variables into appropriate data types
-   Standardised categorical variables
-   Created derived underwriting metrics

::: notes
The data preparation stage improved analytical quality and ensured consistency across the portfolio dataset. Several derived variables were constructed to support underwriting analysis.
:::

## Derived Variables

| Variable           | Purpose                                |
|--------------------|----------------------------------------|
| Loss Ratio         | Measures underwriting profitability    |
| Profitability Band | Segments underwriting performance      |
| Age Group          | Groups drivers by demographic category |

::: notes
Loss Ratio - measures the relationship between claims incurred and premium earned. Lower loss ratio indicates stronger profitability and value greater than 1 imply claims incurred exceed premium income.

Profitability Band - categorise policies into interpretable underwriting performance group. (\< 0.6 Profitable), (0.6 – 1.0 Marginal), (\> 1.0 Unprofitable)

Age Group - simplify demographic analysis by grouping drivers into meaningful underwriting categories. (\< 30 Young), (30 – 60 Middle), (\> 60 Senior)
:::

## Distribution of Total Premium

```{r}
ggplot(
  insurance,
  aes(total_premium)
) +

geom_histogram(
  bins = 50,
  fill = "steelblue",
  color = "white"
) +

scale_y_continuous(
  labels = scales::comma
) +

scale_x_continuous(
  labels = scales::dollar,
  breaks = seq(
    0,
    2000,
    by = 500
  )
) +
  
coord_cartesian(
  xlim = c(0, 2000)
) +

labs(
  title = "Distribution of Total Premium",
  x = "Total Premium",
  y = "Policy Count"
) +

theme_minimal()
```

::: notes
Key Insights

-   The premium distribution is highly right skewed, with most policies concentrated within lower premium ranges.
-   A small number of high premium policies contribute disproportionately to total premium revenue, indicating portfolio heterogeneity.
-   The concentration of lower premium policies suggests strong exposure to mass market insurance segments and potentially competitive pricing conditions.
:::

## Claims Incurred Distribution

```{r}
ggplot(
  insurance,
  aes(x = total_incurred)
) +

geom_density(
  fill = "firebrick",
  alpha = 0.6
) +

scale_x_continuous(
  labels = scales::dollar,
  breaks = seq(
    0,
    5000,
    by = 500
  )
) +

coord_cartesian(
  xlim = c(0, 5000)
) +

labs(
  title = "Claims Incurred Distribution",
  x = "Claims Incurred",
  y = "Density"
) +

theme_minimal()
```

::: notes
Key Insights

-   Most claims are concentrated within lower incurred claim values, indicating that minor claims dominate the portfolio.
-   The long right tail of the distribution highlights the presence of infrequent but severe high cost claims contributing to portfolio volatility.
-   The significant variability in claims severity suggests the importance of effective underwriting, reserving, and catastrophe risk management strategies.
:::

## Profitability Segmentation

```{r}
insurance |>
  
  mutate(
    profitability_band = factor(
      profitability_band,
      
      levels = c(
        "Unprofitable",
        "Profitable",
        "Marginal"
      )
    )
  ) |>
  
  ggplot(
    aes(
      x = profitability_band
    )
  ) +
  
  geom_bar(
    aes(
      y = after_stat(
        count / sum(count)
      )
    ),
    
    fill = "steelblue"
  ) +
  
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, 1)
  ) +
  
  labs(
    title = "Profitability Segmentation",
    
    x = "Profitability Band",
    
    y = "Percentage of Policies"
  ) +
  
  theme_minimal()
```

::: notes
Key Insights

-   The portfolio is primarily concentrated within the profitable segment, indicating generally effective underwriting and pricing performance.
-   A smaller proportion of policies fall within the unprofitable category, suggesting the presence of higher risk segments contributing disproportionately to claims costs.
-   The marginal segment highlights policies operating close to break even levels, indicating potential sensitivity to future claims volatility and profitability deterioration.
:::

## Distribution of Driver Age Group

```{r}
insurance |>
  
  mutate(
    age_band = factor(
      age_band,
      
      levels = c(
        "Young",
        "Middle",
        "Senior"
      )
    )
  ) |>
  
  ggplot(
    aes(age_band)
  ) +
  
  geom_bar(
    aes(
      y = after_stat(
        count / sum(count)
      )
    ),
    
    fill = "steelblue"
  ) +
  
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, 1)
  ) +
  
  labs(
    title = "Distribution of Driver Age Group",
    
    x = "Age Group",
    
    y = "Percentage of Policies"
  ) +
  
  theme_minimal()
```

::: notes
Key Insights

-   The portfolio is primarily concentrated within the middle aged driver segment, suggesting stronger exposure to relatively stable and experienced drivers.
-   Young drivers represent a smaller proportion of the portfolio but may contribute disproportionately to underwriting risk and claims frequency.
-   Senior drivers form the smallest demographic segment, highlighting comparatively lower portfolio exposure to ageing driver risk factors.
:::

## Distribution of Policy Type

```{r}
insurance |>
  mutate(
    policy_type = factor(
      policy_type,
      levels = c(
        "CC",
        "COMP_E",
        "TPG",
        "COMP_N",
        "TP"
      )
    )
  ) |>

  ggplot(
    aes(
      x = policy_type,
      fill = policy_type
    )
  ) +

  geom_bar() +

  scale_fill_manual(
    values = c(
      "CC" = "#7FC97F",
      "COMP_E" = "#80B1D3",
      "TPG" = "#F8766D",
      "COMP_N" = "#B3DE69",
      "TP" = "#FDB462"
    ),

    breaks = c(
      "CC",
      "COMP_E",
      "TPG",
      "COMP_N",
      "TP"
    ),

    labels = c(
      "CC" = "CC: Third-party liability combined",
      "COMP_E" = "COMP_E: Comprehensive with excess",
      "TPG" = "TPG: Third-party liability with glass coverage",
      "COMP_N" = "COMP_N: Comprehensive without excess",
      "TP" = "TP: Basic third-party liability coverage only"
    )
  ) +

  scale_y_continuous(
    labels = scales::comma
  ) +

  labs(
    title = "Distribution of Policy Type",
    x = "Policy Type",
    y = "Policy Count",
    fill = "Policy Description"
  ) +

  theme_minimal()
```

::: notes
Key Insights

-   The portfolio is concentrated within selected policy types, indicating stronger customer demand and underwriting focus in specific insurance products.
-   Comprehensive and third-party coverage products exhibit differing policy concentrations, reflecting variation in customer protection preferences and pricing considerations.
-   The presence of multiple policy categories demonstrates portfolio diversification, which supports underwriting balance and reduces dependence on a single product segment.
:::

## Vehicle Value vs Claims Incurred

```{r}
ggplot(
  insurance,
  aes(
    x = vehicle_value,
    y = total_incurred
  )
) +

geom_point(
  alpha = 0.5,
  color = "steelblue"
) +

scale_y_continuous(
  labels = scales::dollar,
  breaks = seq(
    0,
    200000,
    by = 50000
  )
) +

scale_x_continuous(
  labels = scales::dollar,
  breaks = seq(
    0,
    200000,
    by = 50000
  )
) +

coord_cartesian(
  xlim = c(0, 200000),
  ylim = c(0, 200000)
) +

labs(
  title = "Vehicle Value vs Claims Incurred",
  x = "Vehicle Value",
  y = "Total Incurred Claims"
) +

theme_minimal()
```

::: notes
Key Insights

-   Higher value vehicles generally tend to produce larger incurred claims, indicating a positive association between vehicle value and claim severity.
-   Most observations are concentrated within lower vehicle values and moderate claims ranges, suggesting strong exposure to standard market segments.
-   Several extreme claim observations highlight the presence of high severity loss events contributing disproportionately to portfolio volatility and underwriting risk.
:::

## Profitability Segmentation by Policy Type and Year

```{r}
insurance |>
  
  mutate(
    
    policy_type = factor(
      policy_type,
      levels = c(
        "CC",
        "COMP_E",
        "TPG",
        "COMP_N",
        "TP"
      )
    ),
    
    profitability_band = factor(
      profitability_band,
      levels = c(
        "Unprofitable",
        "Profitable",
        "Marginal"
      )
    )
    
  ) |>
  
  count(
    year,
    policy_type,
    profitability_band
  ) |>
  
  ggplot(
    aes(
      x = factor(year),
      y = policy_type,
      fill = n
    )
  ) +

  geom_tile(
    color = "white"
  ) +

  facet_wrap(
    ~ profitability_band
  ) +

  scale_fill_continuous_sequential(
    
    palette = "BluGrn",

    guide = guide_colorbar(
      barwidth = 20,
      barheight = 1.5,
      title.position = "top",
      title.hjust = 0.5
    )
  ) +

  labs(
    title = "Profitability Segmentation by Policy Type and Year",
    x = "Year",
    y = "Policy Type",
    fill = "Number of Policies"
  ) +

  theme_minimal() +
  
  theme(
    legend.position = "bottom"
  )
```

::: notes
Key Insights

-   Certain policy types exhibit stronger concentration within profitable segments, suggesting more effective underwriting and pricing performance.
-   Variations in colour intensity across years reveal changing portfolio composition and profitability trends over time, supporting temporal underwriting evaluation.
-   The heatmap reveals portfolio imbalance across policy types, suggesting that certain insurance products dominate the insurer’s business mix and may contribute disproportionately to overall portfolio risk.
:::

## Portfolio Composition and Profitability by Fuel Type

```{r}
fuel_summary <- insurance |>
  
  group_by(fuel_type) |>
  
  summarise(
    
    policy_count = n(),
    
    policy_percentage = n() / nrow(insurance),
    
    avg_loss_ratio = mean(
      loss_ratio,
      na.rm = TRUE
    )
    
  )

ggplot(
  fuel_summary,
  
  aes(
    x = reorder(
      fuel_type,
      policy_percentage
    )
  )
) +

geom_col(
  
  aes(
    y = policy_percentage
  ),
  
  fill = "steelblue",
  alpha = 0.8
) +

scale_x_discrete(
  
  labels = c(
    D = "Diesel",
    G = "Gasoline"
  )
) +

geom_line(
  
  aes(
    y = avg_loss_ratio,
    group = 1
  ),
  
  color = "firebrick",
  linewidth = 1.2
) +

geom_point(
  
  aes(
    y = avg_loss_ratio
  ),
  
  color = "firebrick",
  size = 3
) +

scale_y_continuous(
  
  name = "Percentage of Policies",
  
  labels = scales::percent,
  
  limits = c(0, 1),
  
  sec.axis = sec_axis(
    
    ~ .,
    
    labels = scales::percent,
    
    name = "Average Loss Ratio"
  )
) +

labs(
  title = "Portfolio Composition and Profitability by Fuel Type",
  x = "Fuel Type"
) +

theme_minimal()
```

::: notes
Key Insights

-   Diesel dominates the portfolio composition, indicating stronger underwriting exposure and customer concentration within that vehicle segment.
-   Differences in average loss ratio across fuel types suggest that underwriting profitability varies between diesel and gasoline vehicles.
-   Fuel types with both high portfolio concentration and elevated loss ratios may represent important areas of underwriting risk requiring closer monitoring and pricing refinement.
:::

## Driver Age Distribution Across Claims Risk

```{r}
insurance <- insurance |>
  mutate(
    bonus_score = factor(
      bonus_score,
      levels = c("G", "N", "B"),
      labels = c(
        G = "Good Claims",
        N = "Neutral Claims",
        B = "Bad Claims"
      )
    )
  )

ridge_plot <- insurance |>
  ggplot(
    aes(
      x = driver_age,
      y = bonus_score,
      fill = bonus_score
    )
  ) +
  geom_density_ridges(
    alpha = 0.7,
    color = "black",
    scale = 1.2
  ) +
  scale_fill_manual(
    values = c(
      "#7FC97F",
      "#80B1D3",
      "#F8766D"
    )
  ) +
  
  scale_x_continuous(
    
    breaks = c(
      20,
      40,
      60,
      80
    ),
    
    limits = c(
      15,
      95
    )
  
  ) +
  
  labs(
    title = "Driver Age Distribution Across Claims Risk",
    x = NULL,
    y = NULL,
    fill = "Bonus Score"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none"
  )

box_plot <- insurance |>
  ggplot(
    aes(
      x = driver_age,
      y = bonus_score,
      fill = bonus_score
    )
  ) +
  geom_boxplot(
    alpha = 0.8,
    outlier.alpha = 0.5
  ) +
  stat_summary(
    fun = median,
    geom = "label",
    aes(label = round(after_stat(x), 1)),
    fill = "white",
    color = "red",
    size = 3
  ) +
  scale_fill_manual(
    values = c(
      "#7FC97F",
      "#80B1D3",
      "#F8766D"
    )
  ) +
  
  scale_x_continuous(
    
    breaks = c(
      20,
      40,
      60,
      80
    ),
    
    limits = c(
      15,
      95
    )
    
  ) +
  
  labs(
    x = "Driver Age",
    y = NULL,
    fill = "Bonus Score"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
  )

ridge_plot / box_plot
```

::: notes
Key Insights

-   Drivers with good claims history appear more concentrated within middle-age groups, suggesting experience may contribute to lower insurance risk.
-   The bad claims segment exhibits greater variability in driver age distribution, indicating that higher risk claims behaviour spans multiple demographic groups.
-   Differences in median driver age across claims categories suggest that age may be an important factor influencing underwriting performance and claims experience.
:::

## Conclusion

1.  Visual analytics successfully identified key profitability and risk drivers.
2.  Portfolio risk and profitability are unevenly distributed.
3.  Demographic characteristics significantly influence claims behaviour.
4.  Advanced visual analytics techniques improved analytical interpretation and storytelling
5.  The analysis provides strong business value for insurers.
